---
title: "Partial reproduction of the LDA models and predictions from Hernández Fernández 2001 & 2003"
author: "Ben Marwick and Gillian Wong"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
 bookdown::html_document2:
   keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      fig.path = "README_figs/README-")
```



### Hernández Fernández (2001) and the whole mammal bioclimatic model {-#whole}

Hernández Fernández (2001, 2003) proposed a bioclimatic model for palaeoclimatic reconstruction based on the assumption of a correlation between climate and mammal community composition. 

In his 2001 paper we see table 2 which has data for whole mammal faunal bioclimatic spectra of the localities. Each column represents a bioclimatic component, and BCi values. From Appendix 1 of the 2001 paper we can get the actual climatic zones for the localities in table 2. Using these two tables we can make a labelled training set to train a machine learning model that can predict the climate zone of new faunal assemblages. 

```{r fh-2001-data}
library(docxtractr)
library(tidyverse)
# get the BCi and Cl. values from https://onlinelibrary.wiley.com/doi/full/10.1046/j.1466-822x.2001.00218.x
# Table 2 Whole mammal faunal bioclimatic spectra of the localities. Each column 
# represents a bioclimatic component (see Table 1), and BCi values 
# Appendix 1 Characteristics of studied localities (Cl., climate zone)

fh_2001_tbl_2 <- read_csv(here::here("data/Fernández-2001-table-2.csv"))
fh_2001_tbl_appx_1 <- read_csv(here::here("data/Fernández-2001-table-appendix1.csv"))

fh_2003_tbl_appx_1.1 <- # Whole mammal faunal bioclimatic spectra, same as fh_2001_tbl_2
  docx_extract_tbl(read_docx(here::here("data/geb_057_sm_appendix1.1.doc")))

fh_2001_BCi_values_and_Cl <- 
  left_join(fh_2001_tbl_2, 
            fh_2001_tbl_appx_1) %>% 
  dplyr::select(-c(
                   Latitude,  
                   Longitude, 
                   Altitude,
                   References,
                   Country,
                   N))

library(knitr)
kable(fh_2001_BCi_values_and_Cl, caption = "Whole mammal faunal bioclimatic spectra from 50 locations")
```

Using these data, we can roughly follow Hernández Fernández's method to compute a Linear Discriminant Analysis to generate a model to predict the climate zone of a new faunal assemblage:

```{r fh-model1}
library(caret)
set.seed(998)

fh_2001_BCi_values_and_Cl <- select(fh_2001_BCi_values_and_Cl, -Locality)

inTraining1 <- createDataPartition(fh_2001_BCi_values_and_Cl$Cl., 
                                  p = .75, 
                                  list = FALSE)

training1 <- fh_2001_BCi_values_and_Cl[ inTraining1,]
testing1  <- fh_2001_BCi_values_and_Cl[-inTraining1,]

# estimate pre processing parameters
# center and scale
preProcValues1 <- preProcess(training1, method = c("center", "scale"))

trainTransformed1 <-  
 predict(preProcValues1, training1)

testTransformed1 <- 
 predict(preProcValues1, testing1)

# now run DFA using Linear Discriminant analysis (LDA) on the normalized data 

# Basic Parameter Tuning
fitControl <- trainControl(## 10-fold CV
  method = "repeatedcv",
  number = 10,
  ## repeated ten times
  repeats = 10)

# fit the model to the training data
ldaFit1 <- train(Cl. ~ ., 
                 trControl = fitControl,
                 data = trainTransformed1, 
                 method = "lda")
```

We can inspect the performance of this model on the training set, and we can see one location mis-classified 

```{r fh-model1-training}
# model against the training set 
lda_train <- predict(MASS::lda(Cl. ~ ., data = trainTransformed1))
trainTransformed1$lda <- lda_train$class
kable(table(trainTransformed1$lda, trainTransformed1$Cl.),
      caption = "Model performance with training data")
```

And we can inspect the model on the testing data set, and we see less accuracy:

```{r fh-model1-testing}
# use this model to predict the classifications of the test data
test_predictions1 <- 
  predict(ldaFit1, 
          newdata = testTransformed1)
# model against the testing set 
kable(table(test_predictions1, testTransformed1$Cl.),
      caption = "Model performance with hold-out test data")

pred.accuracy = round(mean(test_predictions1 == testTransformed1$Cl.) * 100, 2)
```

And we find a prediction accuracy of `r pred.accuracy` % in this case, not bad. 

In Appendix 3, Hernández Fernández (2001) shows the results of his LDA model applied to the locations in his training data set. He appears to have tested his model on the same data set that he used to train it, which is not ideal. Perhaps not suprisingly, there is a very good fit, and he claims a 98% correct classification rate. Only one location was incorrect, the same as we observed with our training data. From these results we can conclude that we have successfully reproduced some key results of Hernández Fernández (2001). 

### Hernández Fernández (2003) and whole fauna biospectra at 13 new locations {-#new}

In his 2003 paper, Hernández Fernández attempts to infer past climatic conditions using mammal fossil associations as source data. In this 2003 paper we also have 13 new recent faunas presented for testing. So let's attempt to reproduce Hernández Fernández's results with these 13 new fauna samples. We will train a LDA model on all 50 locations from the 2001 paper. Then we will use this model to classify the climate zones of these 13 new locations in the 2003 paper. Then we'll check to see how accurate the model was with these 13 new locations. This is a direct reproduction of the key results in the 2003 paper. 

Here are the 13 new locations from appendix 3.1 in the 2003 paper that we will use as the test dataset, just as Hernández Fernández does:

```{r fh-model2-import-data}
hf_2003_appdx_3.1 <- 
  read_docx(here::here("data/geb_057_sm_appendix 3.1.doc")) %>% 
  docx_extract_tbl() 

# get the table from the appendix word doc
hf_2003_appdx_3.1_tbl <-
  hf_2003_appdx_3.1 %>%
  mutate_at(vars(2:11), ~ parse_number(.)) %>%
  dplyr::select(-Whole.mammal) %>%
  set_names(c('I', 'II', 'II/III', 'III', 'IV', 'V',  'VI',  'VII', 'VIII', 'IX'))

kable(hf_2003_appdx_3.1, caption = "Whole mammal biospectra from 13 new locations")
```

Here is the LDA model that we generate from all 50 locations in the 2001 paper.  Hernández Fernández used SPSS, so we don't know the exact configuration of his model. But we can see elshwere online that R and SPSS can give good agreement with LDA.

```{r fh-model2-lda}
library(caret)
set.seed(998)

training2 <- fh_2001_BCi_values_and_Cl

# estimate pre processing parameters
# center and scale
preProcValues2 <- preProcess(training2, method = c("center", "scale"))

trainTransformed2 <-  
 predict(preProcValues2, training2)

# now run DFA using Linear Discriminant analysis (LDA) on the normalized data 

# Basic Parameter Tuning
fitControl <- trainControl(## 10-fold CV
  method = "repeatedcv",
  number = 10,
  ## repeated ten times
  repeats = 10)

# fit the model to the training data
ldaFit2 <- train(Cl. ~ ., 
                 trControl = fitControl,
                 data = trainTransformed2, 
                 method = "lda")
```

Table 2 in the 2003 paper shows the actual Climate zones for the 13 new locations, so let's import that here so we can compare with our model's predictions:

```{r fh-model2-import-more}
fh_2003_tbl_2 <- read_csv(here::here("data/Fernández-2003-table-2.csv"))
fh_2003_tbl_2_no_na <- filter(fh_2003_tbl_2, !is.na(`Climate zone`))

kable(fh_2003_tbl_2_no_na, caption = "Actual climate zones of the 13 new locations")
```

Now we can inspect the model on the testing data set of the 13 new locations from the 2003 paper. Below we can see a pretty accurate classification from our model's predictions and the actual climate zone reported in Hernández Fernández (2003) table 2:

```{r fh-model2-compare}
# use this model to predict the classifications of the test data on the 13 new locations
testing2  <- hf_2003_appdx_3.1_tbl

testTransformed2 <- 
 predict(preProcValues2, testing2)

test_predictions_ldaFit2 <- 
  predict(ldaFit2, 
          newdata = testTransformed2)

# model against the testing set 
kable(tibble(Location = hf_2003_appdx_3.1$Whole.mammal, 
       `Predicted by our model`= test_predictions_ldaFit2,
       `Actual zone from table 2` = fh_2003_tbl_2_no_na$`Climate zone`),
      caption = "Model predictions versus actual climate zones of the 13 new locations")

```

### Hernández Fernández (2003) and Rodentia data {-#rodent}

Let's take a look at how effective this model is with the Rodent data. In his 2003 paper we see 'Appendix 1.2 Rodentia faunal bioclimatic spectra of the localities' which appears to be model training data for predicting climate zones from rodent data only. The rodent data of 13 new locations are given in 'Appendix 3.2 Rodentia faunal bioclimatic spectra of the 13 new localities used for the validation of the bioclimatic model.' So let's see if we can reproduce the rodent analysis. We'll generate a new LDA model using the rodent data from 50 locations as the training set, then predict the locations of the 13 new faunas using only their rodent data. 

```{r fh-model3-import}
hf_2003_appdx_1.2 <- # training set
  read_docx(here::here("data/geb_057_sm_appendix1.2.doc")) %>% 
  docx_extract_tbl() 

hf_2003_appdx_2.2 <- # coefficients 
  read_docx(here::here("data/geb_057_sm_appendix 2.2.doc")) %>% 
  docx_extract_tbl() 

hf_2003_appdx_3.2 <- # 13 new locations
  read_docx(here::here("data/geb_057_sm_appendix 3.2.doc")) %>% 
  docx_extract_tbl() 

kable(hf_2003_appdx_1.2, caption="Rodent data from 50 locations")
kable(hf_2003_appdx_3.2, caption="Rodent data from 13 locations")
```

Here we generate a LDA model based on the rodent data from 50 locations, as we see in Appendix 1.2 of the 2003 paper: 

```{r fh-model3-lda}
library(caret)
set.seed(998)

# get the actual climate zones, omit NA, and covert chr to dbl
fh_2003_rodent_BCi_values_and_Cl <- 
  left_join(hf_2003_appdx_1.2, 
            fh_2001_tbl_appx_1) %>% 
  dplyr::select(-c(Locality , 
                   Latitude,  
                   Longitude, 
                   Altitude,
                   References,
                   Country,
                   N)) %>% 
  filter(!is.na(Cl.)) %>% 
  mutate_at(vars(1:10), ~parse_double(.))

training3 <- fh_2003_rodent_BCi_values_and_Cl

# estimate pre processing parameters
# center and scale
preProcValues3 <- preProcess(training3, method = c("center", "scale"))

trainTransformed3 <-  
 predict(preProcValues3, training3)

# now run DFA using Linear Discriminant analysis (LDA) on the normalized data 

# Basic Parameter Tuning
fitControl <- trainControl(## 10-fold CV
  method = "repeatedcv",
  number = 10,
  ## repeated ten times
  repeats = 10)

# fit the model to the training data
ldaFit3 <- train(Cl. ~ ., 
                 trControl = fitControl,
                 data = trainTransformed3, 
                 method = "lda")
```

Now we use this rodent LDA model to classify the 13 new locations presented in Appendix 3.2 of the 2003 paper. Once again we see a good classification, all the predictions match up with the actual zones. 

```{r fh-model3-predict}
# use this model to predict the classifications of the test data on the 13 new locations
testing3  <- 
  hf_2003_appdx_3.2 %>% 
  select(-Rodentia) %>% 
  mutate_at(vars(1:10), ~parse_double(.))

testTransformed3 <- 
 predict(preProcValues3, testing3)

test_predictions_ldaFit3 <- 
  predict(ldaFit3, 
          newdata = testTransformed3)

# model against the testing set 
kable(tibble(Location = hf_2003_appdx_3.2$Rodentia, 
       `Predicted by our model`= test_predictions_ldaFit3,
       `Actual zone from table 2` = fh_2003_tbl_2_no_na$`Climate zone`), 
      caption = "Model predictions versus actual climate zones for Rodent data from the 13 new locations")

```

The coefficients of Hernández Fernández (2003) rodent-only model are presented in 'Appendix 2.2 Coefficients for discriminant functions calculated from rodent faunal bioclimatic components and centroids, for each discriminant function, for every locality's biome group.' We can also show the coefficients from our rodent model. 

Here are Hernández Fernández (2003) rodent LDA coefficients:

```{r fh-coef}
kable(hf_2003_appdx_2.2, caption = "Hernández Fernández (2003) rodent-only coefficients")
```

And here are our rodent LDA coefficients, they are very different, and it's not clear why. I can see comparisons of SPSS and R LDA methods online and they can get the same coefficients to within 2 decimal places. 

```{r our-coef}
kable(ldaFit3$finalModel$scaling, , caption = "Our rodent-only coefficients")
```

Let's try it again without any pre-processing of the data, still our coefficients are not similar to those published by Hernández Fernández (2003)

```{r}

# now run DFA using Linear Discriminant analysis (LDA) on the normalized data 

# Basic Parameter Tuning
fitControl <- trainControl(## 10-fold CV
  method = "repeatedcv",
  number = 10,
  ## repeated ten times
  repeats = 10)

# fit the model to the training data
ldaFit4 <- train(Cl. ~ ., 
                 trControl = fitControl,
                 data = training3, 
                 method = "lda")

# predict Climate zones of new data

test_predictions_ldaFit4 <- 
  predict(ldaFit4, 
          newdata = testing3)

kable(ldaFit4$finalModel$scaling, , caption = "Our rodent-only coefficients, no transformation to model inputs")
```



#### References {-#refs}

Fernández, M. H. (2001). Bioclimatic discriminant capacity of terrestrial mammal faunas. Global Ecology and Biogeography, 10(2), 189-204. <https://doi.org/10.1046/j.1466-822x.2001.00218.x>

Fernández, M. H., & Peláez‐Campomanes, P. (2003). The bioclimatic model: a method of palaeoclimatic qualitative inference based on mammal associations. _Global Ecology and Biogeography_, 12(6), 507-517. <https://doi.org/10.1046/j.1466-822X.2003.00057.x>

```{r}
sessioninfo::session_info()
```

